---
title: "Dicrurus Colorspace"
output: github_document
---

Occurrences come from a georeferenced specimen database and from the Global Biodiversity Informatics Facility, specifically: GBIF.org (10 March 2022) GBIF Occurrence Download https://doi.org/10.15468/dl.b5q394 

```{r,echo=F}
rm(list=ls())
filepath="~/Dropbox/Manuscripts/Dicrurus/"
gis="~/Dropbox/GIS/"
```
```{r}
library(data.table)
library(factoextra)
library(fossil) # Rarefy code
library(maptools)
library(MASS)
library(raster)
library(rnaturalearth)
library(rnaturalearthhires)
library(rgdal)
library(sf)
library(tidyverse)
library(vegan)
```

# Specimen Localities

Import coordinates for colorspace specimens from KMZ file.

```{r}
# converted from KMZ in QGIS

ocean=readOGR(paste0(gis,"ne_10m_ocean/ne_10m_ocean.shp"))
lakes=readOGR(paste0(gis,"ne_10m_lakes/ne_10m_lakes.shp"))
elev=raster(paste0(gis,"elevation_1KMmd_GMTEDmd.tif"))
specimen.xy=read_csv(paste0(filepath,'dicrurus_locs_mar2022.csv'))

# NOTE that long and lat were switched during the conversion
ylim=c(min(specimen.xy$Long),max(specimen.xy$Long))
xlim=c(min(specimen.xy$Lat),max(specimen.xy$Lat))
```

Correct the species field.

```{r}
unique(specimen.xy$species)
```

```{r}
# ORDER MATTERS! Most complex first.
specimen.xy$species=gsub("divar(east)/ads.jubaensis?","Dicrurus divaricatus/adsimilis jubaensis",specimen.xy$species)
specimen.xy$species=gsub("ads.jubaensis/ads.fugax?","Dicrurus adsimilis jubaensis/fugax",specimen.xy$species)
specimen.xy$species=gsub("divar(east)/divar(west)?","Dicrurus divaricatus east/west",specimen.xy$species)
specimen.xy$species=gsub("ads.adsimilis/ads.fugax?","Dicrurus adsimilis adsimilis/fugax",specimen.xy$species)
specimen.xy$species=gsub("ads.fugax","Dicrurus adsimilis fugax",specimen.xy$species)
specimen.xy$species=gsub("apivorus","Dicrurus aapivorus",specimen.xy$species)
specimen.xy$species=gsub("ads.jubaensis","Dicrurus adsimilis jubaensis",specimen.xy$species)
specimen.xy$species=gsub("mod.coracinus","Dicrurus modestus coracinus",specimen.xy$species)
specimen.xy$species=gsub("divar(west)","Dicrurus divaricatus west",specimen.xy$species)
specimen.xy$species=gsub("atactus","Dicrurus atactus",specimen.xy$species)
specimen.xy$species=gsub("ads.adsimilis","Dicrurus adsimilis adsimilis",specimen.xy$species)
specimen.xy$species=gsub("divar(east)","Dicrurus divaricatus east",specimen.xy$species)
specimen.xy$species=gsub("ads.fugax(usambara)","Dicrurus adsimilis fugax",specimen.xy$species)
specimen.xy$species=gsub("forficatus","Dicrurus forficatus",specimen.xy$species)
```


```{r}
plot(elev,xlim=xlim*1.1,ylim=ylim*1.1)
plot(ocean,col="#A6BBFF",add=T)
plot(lakes,col="#A6BBFF",add=T)
points(x=specimen.xy$Lat,y=specimen.xy$Long,
       col=as.factor(specimen.xy$species),pch=19)
```

Link the occurrences with specimen metadata. This has to be done based on placenames.

```{r,eval=F}
colnames(specimen.xy)[4:5]=c("Lat","Long")
specimen.data=specimen.xy%>%
  dplyr::select(Long,Lat,species)%>%
  unique()
colnames(specimen.data)[3]="Species"

# removing location linking step from previous version
```

## Merge with GBIF

We are going to merge the above with GBIF data.

```{r,eval=F}
# read GBIF
gbif=read.delim(paste0(filepath,"dicrurus_gbif.csv"),
             sep="\t") # specify tab separation

# remove extraneous columns
# parse to relevant columns
gbif=gbif%>%
  # remove gridded data as determined in QGIS
  filter(institutionCode!='6992',
         institutionCode!='6993',
         institutionCode!='APLORI',
         institutionCode!='FIAO')%>%
  dplyr::select(genus,species,infraspecificEpithet,
         scientificName,individualCount,
         decimalLongitude,decimalLatitude,coordinateUncertaintyInMeters,
         coordinatePrecision,
         day,month,year,countryCode)

# write reduced file
write_csv(gbif,file=paste0(filepath,"reduced_gbif.csv"))

# remove NA vals from uncertatainty
# assume accurate unless states otherwise
gbif[is.na(gbif$coordinateUncertaintyInMeters),
  "coordinateUncertaintyInMeters"]=0

# localities only
gbif=gbif%>%
  # remove uncertainty over 10km
  filter(coordinateUncertaintyInMeters<=10000)%>%
  select(species,infraspecificEpithet,
         decimalLongitude,decimalLatitude,
         countryCode,month)%>%
  # get unique
  unique()

# convert specimen sheet to same format
#colnames(gbif)
#colnames(specimen.data)
specimen.data$month=NA
specimen.data$infraspecificEpithet=NA
specimen.data$countryCode=NA

colnames(specimen.data)[1:3]=c("decimalLongitude","decimalLatitude","species")

col.order=colnames(gbif)
specimen.data=specimen.data%>%
  dplyr::select(paste0(col.order))
```
```{r,eval=F}
sp.field=unique(specimen.data$species)

# adsimilis fugax
i=1
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]="fugax"
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus adsimilis"

# Dicrurus aapivorus
i=2
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]="apivorus"
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus adsimilis"

# Dicrurus adsimilis jubaensis
i=3
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]="jubaensis"
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus adsimilis"

# Dicrurus adsimilis/divaricatus
i=4
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]=NA
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus adsimilis/divaricatus"

# Dicrurus adsimilis jubaensis/fugax
i=5
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]='jubaensis/fugax'
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus adsimilis"

# Dicrurus modestus coracinus
i=6
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]='coracinus'
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus modestus"

# Dicrurus divaricatus, ssp unknown
i=7
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]=NA
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus divaricatus"

# Dicrurus divaricatus divaricatus
i=8
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]='divaricatus'
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus divaricatus"

# Dicrurus atactus
i=9
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]=NA
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus atactus"

# Dicrurus adsimilis adsimilis/fugax
i=10
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]='adsimilis/fugax'
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus adsimilis"

# Dicrurus adsimilis adsimilis
i=11
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]='adsimilis'
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus adsimilis"

# Dicrurus divaricatus lugubris
i=12
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]='lugubris'
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus divaricatus"

# Dicrurus adsimilis fugax(usambara)
i=13
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]='usambarae'
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus adsimilis"

# Dicrurus forficatus forficatus
i=14
sp.field[i]
specimen.data$infraspecificEpithet[specimen.data$species==sp.field[i]]='forficatus'
specimen.data$species[specimen.data$species==sp.field[i]]="Dicrurus forficatus"

write_csv(specimen.data,file=paste0("dicrurus_specimen_localities_fixed.csv"))
```
```{r,eval=F}
write_csv(gbif,file = paste0(filepath,"dicrurus_gbif_localities.csv"))
```
```{r}
gbif=read_csv(paste0(filepath,"dicrurus_gbif_localities.csv"))
specimen.data=read_csv(paste0("dicrurus_specimen_localities_fixed.csv"))
```
```{r}
plot(elev,xlim=xlim*1.1,ylim=ylim*1.1)
plot(ocean,col="#A6BBFF",add=T)
plot(lakes,col="#A6BBFF",add=T)
points(specimen.xy$Lat,specimen.xy$Long,col="red",pch=19) # coordswitch
points(gbif$decimalLongitude,gbif$decimalLatitude,
       col="black",pch=".")
```

The samples specimens cover a lot more of northeastern Africa than the other data. Note how data-dense Southern Africa is relative to the rest of the continent!

Prepare data for merge.

```{r,eval=F}
# create voucher column
#colnames(specimen.data)==colnames(gbif)
colnames(gbif)=colnames(specimen.data)=c("SciName","Subspecies","Long","Lat","Country","Month")

# only gbif data for now; to be changed later

gbif=gbif%>%
  dplyr::select(SciName,Subspecies,Long,Lat)
spec=specimen.data%>%
  dplyr::select(SciName,Subspecies,Long,Lat)

gbif$Source="GBIF"
spec$Source="Specimen"

all.dat=rbind(gbif,spec)

write_csv(all.dat,paste0(filepath,"combined_data.csv"))
```

# Population assignation

Ensure that populations are correctly assigned. Many taxa have holdover from past taxonomic treatises. The study is following the phylogeny of Fuchs et al., which has been largely adopted by Birds of the World. Thus, we are correcting taxonomy to Birds of the World 2022. This is where things get tricky! Note that specimens are likely correctly attributed, so we are leaving them alone.

```{r}
data=read_csv(paste0(filepath,"combined_data.csv"))

#data$SciName=as.factor(data$SciName)
#data$Subspecies=as.factor(data$Subspecies)
#summary(data)
```

```{r}
summary(as.factor(data$SciName))
```

Looks like the main study species are present.

```{r}
# plot species in dataset

species.plotter=function(sciname,data){
  data2=data%>%
    filter(SciName==sciname)%>%
    dplyr::select(Long,Lat,Source)
  
  plot(elev,xlim=xlim*1.1,ylim=ylim*1.1)
  plot(ocean,col="#A6BBFF",add=T)
  plot(lakes,col="#A6BBFF",add=T)
  index1=which(data2$Source=="GBIF")
  index2=which(data2$Source=="Specimen")
  points(data2[index1,1:2],col='black',pch=19)
  if(length(index2)>0){points(data2[index2,1:2],col='red',pch=19)}
}
```

```{r}
# load range shapefiles

shps=paste0(filepath,"ssp_shp/",list.files(paste0(filepath,"ssp_shp/"),pattern="*.shp"))
```

```{r}
# remove birds with no species assigned
data=data[-which(is.na(data$SciName)),]
```

### Rarefying the data

MVEs are supposed to be robust to be more robust to data bias, but given issues with thresholding, biases can still remain. Given that some countries are exponentially more sampled that other (e.g., South Africa vs. Chad), we are also going to run a rarefy code. This is a custom code written by Dr. Joe Manthey; it is loaded locally and is not included with this release.

```{r,echo=F}
source("~/Dropbox/Manuscripts/Dicrurus/Manthey_rarefy_2017_dicrurus.R")

rarefy_species=function(species,km){
  subset.data=data%>%
    filter(SciName==species)%>%
    filter(Source=='GBIF')
  specimens=data%>%
    filter(SciName==species)%>%
    filter(Source=='Specimen')
  if(nrow(subset.data)<1){
    print("ERROR: check your spelling.")
    break}
  data=data%>%filter(SciName!=species)
  
  # run modified code
  subset.data=rarefy(input=subset.data,distance.km=km)
  data=rbind(data,subset.data,specimens)
  return(data)
}
```

we are going to rarefy down to 20 km to account for potentially long traveling counts

## *Dicrurus occidentalis*

Firstly, rarefy.

```{r}
data=rarefy_species(species = "Dicrurus occidentalis",km=20)
```

```{r}
which(data$Subspecies%like%"occidentalis")
```
There is nothing ascribed to the subspecies *occidentalis*.

```{r}
species.plotter(data=data,sciname="Dicrurus occidentalis")
```

All individuals are within the range of *D. occidentalis*. There are no specimens.

## *Dicrurus sharpei*

```{r}
data=rarefy_species(species="Dicrurus sharpei",km=20)
```

```{r}
data[which(data$Subspecies%like%"sharpei"),]
```
There are several misclassified *D. ludwigii* still.

```{r}
index=which(data$Subspecies%like%"sharpei")

data[index,"SciName"]="Dicrurus sharpei"
data[index,"Subspecies"]=NA
```

```{r}
species.plotter(data=data,sciname="Dicrurus sharpei")
```

There are lots of erroneous western records, which refer to *D. occidentalis*. There are no specimens. We also are limited to south of the Benue, as this is where the species is certainly known. Note that records from Jos are best omitted from this dataset at this time, and that records west of the Niger are likely *D. occidentalis*.

```{r}
data[which(data$SciName%like%"sharpei"&data$Long<3.45),
     "SciName"]="Dicrurus occidentalis"

# remove north of Benue
data=data[-which(data$SciName%like%"sharpei"&
                   data$Lat>8.9&data$Long<10.2),]
```

## *Dicrurus ludwigii*

```{r}
data=rarefy_species(species="Dicrurus ludwigii",km=20)
```

This species consists of the following subspecies:

1. *saturnus* from Angola to Katanga and Zambia
2. *muenzneri* from Somalia, Kenya, and Tanzania (coastal?)
3. *tephrogaster* from Malawi, Zimbabwi, and Mozambique "north of Limpopo River"
4. *ludwigii* for all localities S of Limpopo R

```{r}
unique(data$Subspecies[data$SciName%like%"ludwigii"])
```
All subspecies appear to be part of the species complex. Strangely, no *saturnus* are labeled in the group.

```{r}
species.plotter(data=data,sciname = "Dicrurus ludwigii")
```

*Note* that populations in W Africa are labeled incorrectly. Using the above threshold:

```{r}
data[which(data$SciName%like%"ludwigii"&data$Long<3.45),
     "SciName"]="Dicrurus occidentalis"
data[which(data$SciName%like%"ludwigii"&
             data$Long<20&data$Lat>0),
     "SciName"]="Dicrurus sharpei"
```

## *Dicrurus atripennis*

```{r}
data=rarefy_species(species="Dicrurus atripennis",km=20)
```

```{r}
unique(data$Subspecies[data$SciName%like%"atripennis"])
```
No subspecies, as expected.

```{r}
species.plotter(sciname = "Dicrurus atripennis",data=data)
```
There are some northerly, almost Sahelian records that are surprising. The only one that we can immediately say is in error is the Kenyan record.

```{r}
data=data[-which(data$SciName%like%"atripennis"&
             data$Long>32.5),]
```

## *Dicrurus adsimilis*

```{r}
data=rarefy_species(species="Dicrurus adsimilis",km=20)
```

```{r}
summary(as.factor(data$Subspecies[data$SciName%like%"adsimilis"]))
```

Most of these are NA; we need to correct them to subspecies.

```{r}
species.plotter("Dicrurus adsimilis",data=data)
```

Looks like a lot of different taxa are mixed together here.

```{r}
adsimilis=shps[shps%like%'adsimilis']

divaricatus=shps[shps%like%'divaricatus']

tst.shp=c(adsimilis,divaricatus)
```

GBIF data includes some specimens, but it's not *our* verified specimens, so subjecting all to reclassification.

```{r}
summary(as.factor(data$SciName))
```


```{r}
spec.dat=data%>%filter(SciName=="Dicrurus adsimilis")%>%
  filter(Source=="Specimen")
data2=data%>%filter(SciName=="Dicrurus adsimilis")%>%
  filter(Source=="GBIF")
data=data%>%filter(SciName!="Dicrurus adsimilis")

# for taxon "adsimilis"
data2$Subspecies=NA

for(i in 1:length(tst.shp)){
  # load shapefile
  ssp=strsplit(tst.shp[i],"_")[[1]][4]
  ssp=strsplit(ssp,"[.]")[[1]][1]
  
  y=readOGR(tst.shp[i])
  # set data2 so we can do over
  y@data[,]=1
  # read locality data2
  xy=data2%>%dplyr::select(Long,Lat)
  coordinates(xy)= ~ Long + Lat
  crs(xy)=crs(y)
  
  index1=over(xy,y)
  index1=which(index1==1)
  
  data2$Subspecies[index1]=ssp
}

data2$SciName[which(data2$Subspecies=="divaricatus")]="Dicrurus divaricatus"
data2$SciName[which(data2$Subspecies=="lugubris")]="Dicrurus divaricatus"

data2=na.omit(data2)

data=rbind(data,data2,spec.dat)%>%unique()
```

```{r}
species.plotter("Dicrurus adsimilis",data=data)
```

Much better map for this group.

## *Dicrurus divaricatus*

```{r}
data=rarefy_species(species="Dicrurus divaricatus",km=20)
```

Similar to the above, we need to group these together by subspecies as well. Similar to the above, birds near contact zones are not assigned and are removed.

```{r}
species.plotter(data=data,sciname = "Dicrurus divaricatus")
```

Now to assign these to subspecies as well.

```{r}
tst.shp=shps[which(shps%like%"divaricatus")]

data2=data%>%filter(SciName=="Dicrurus divaricatus")
data=data%>%filter(SciName!="Dicrurus divaricatus")

# for taxon "divaricatus"
data2$Subspecies=NA

for(i in 1:length(tst.shp)){
  # load shapefile
  ssp=strsplit(tst.shp[i],"_")[[1]][4]
  ssp=strsplit(ssp,"[.]")[[1]][1]
  
  y=readOGR(tst.shp[i])
  # set data2 so we can do over
  y@data[,]=1
  # read locality data2
  xy=data2%>%dplyr::select(Long,Lat)
  coordinates(xy)= ~ Long + Lat
  crs(xy)=crs(y)
  
  index1=over(xy,y)
  index1=which(index1==1)
  
  data2$Subspecies[index1]=ssp
}

data2=na.omit(data2)

data=rbind(data,data2)%>%unique()
```

## *Dicrurus atactus*

```{r}
data=rarefy_species(species="Dicrurus atactus",km=20)
```

```{r}
species.plotter(data=data,sciname = "Dicrurus atactus")
```

All these records look correct.

## *Dicrurus modestus*

```{r}
data=rarefy_species(species="Dicrurus modestus",km=20)
```

```{r}
species.plotter(data=data,sciname = "Dicrurus modestus")
```

We also need these to be assigned to population.

```{r}
tst.shp=shps[shps%like%'modestus']
```

```{r}
spec.data=data%>%
  filter(SciName=="Dicrurus modestus")%>%
  filter(Source=="Specimen")
data2=data%>%filter(SciName=="Dicrurus modestus")%>%
  filter(Source=="GBIF")
data=data%>%filter(SciName!="Dicrurus modestus")

# for taxon "divaricatus"
data2$Subspecies=NA

for(i in 1:length(tst.shp)){
  # load shapefile
  ssp=strsplit(tst.shp[i],"_")[[1]][4]
  ssp=strsplit(ssp,"[.]")[[1]][1]
  
  y=readOGR(tst.shp[i])
  # set data2 so we can do over
  y@data[,]=1
  # read locality data2
  xy=data2%>%dplyr::select(Long,Lat)
  coordinates(xy)= ~ Long + Lat
  crs(xy)=crs(y)
  
  index1=over(xy,y)
  index1=which(index1==1)
  
  data2$Subspecies[index1]=ssp
}

data2=na.omit(data2)

data2$SciName[which(data2$Subspecies=="atactus")]="Dicrurus atactus"

data=rbind(data,data2,spec.data)%>%unique()
```

```{r}
species.plotter(data=data,sciname="Dicrurus modestus")
```

## *Dicrurus aldabranus*

Excluded from this study.

```{r}
data=data%>%
  filter(SciName!="Dicrurus aldabranus")
```


## *Dicrurus fuscipennis*

Excluded from this study.

```{r}
data=data%>%
  filter(SciName!="Dicrurus fuscipennis")
```


## *Dicrurus forficatus*

```{r}
data=rarefy_species(species="Dicrurus forficatus",km=20)
```

Restrict to nominate, mainland Madagascar.

```{r}
tst.shp=shps[shps%like%'forficatus']

data2=data%>%filter(SciName=="Dicrurus forficatus")
data=data%>%filter(SciName!="Dicrurus forficatus")

# for taxon "divaricatus"
data2$Subspecies=NA

for(i in 1:length(tst.shp)){
  # load shapefile
  ssp=strsplit(tst.shp[i],"_")[[1]][4]
  ssp=strsplit(ssp,"[.]")[[1]][1]
  
  y=readOGR(tst.shp[i])
  # set data2 so we can do over
  y@data[,]=1
  # read locality data2
  xy=data2%>%dplyr::select(Long,Lat)
  coordinates(xy)= ~ Long + Lat
  crs(xy)=crs(y)
  
  index1=over(xy,y)
  index1=which(index1==1)
  
  data2$Subspecies[index1]=ssp
}

data2=na.omit(data2)

data=rbind(data,data2)%>%unique()
```


```{r}
species.plotter(data=data,sciname = "Dicrurus forficatus")
```

## *Dicrurus waldenii*

Excluded from study.

```{r}
data=data%>%
  filter(SciName!="Dicrurus waldenii")
```

## Save data

```{r}
colnames(data)=c('Species','Subspecies',
                 'Long','Lat','Source')
data=data%>%unique()
write_csv(data,paste0(filepath,"reassigned_data.csv"))
```

# Extract Environmental Data

```{r}
data=read_csv(paste0(filepath,"reassigned_data.csv"))
```

First, a quick preview of distributions:

```{r}
plot(ocean,col="#A6BBFF",xlim=xlim*1.1,ylim=ylim*1.1)
points(data[,c('Long','Lat')],col=as.factor(data$Species),pch=20)
```

```{r}
env_layers=paste0(gis,"ENVIREM_30arcsec/",
  list.files(paste0(gis,"ENVIREM_30arcsec/"),pattern="*.bil"))

# remove set count files
env_layers=env_layers[-which(env_layers%like%"growing")]
env_layers=env_layers[-which(env_layers%like%"Count")]
```

```{r,eval=F}
y=stack(env_layers)

coords=data%>%
  dplyr::select(Long,Lat)

env_vals=raster::extract(x=y,y=coords)

data=cbind(data,env_vals)

write_csv(data,paste0(filepath,"env_extracts.csv"))
```

